Inducing Domain Specific Languages for Bayesian Program Learning

### 2018-02-02 paper.tex
edits

The Helmholtz machine~\cite{dayan1995helmholtz} introduced the idea of alternating between
learning a generative model and training a recognition model on samples from the generative model.
We apply this same idea to the problem of program induction,
and so we call our algorithm the \system.
Figure~\ref{feeding} diagrams how these three steps feed off of each other.
\begin{figure}[b]\centering
  \begin{tikzpicture}
    \begin{scope}[shift = {(1,-1)}]
    \node[align = center](DSL) at (6,4) {DSL: $\mathcal{D}$};
    \node[align = center](synthesis) at (4,2) {Search for \\programs: $p$};
    \node[ align = center](recognitionModel) at (6,0) {Recognition \\model: $q$};

@@

\begin{algorithm}[tb]
   \caption{Generative model over programs}
   \label{programGenerativeModel}
   \begin{algorithmic}
     \STATE \textbf{function} sampleProgramFromDSL$(\mathcal{D}, \theta, \tau)$:
  \STATE {\bfseries Input:} DSL $\mathcal{D}$, weight vector $\theta$, type $\tau$
  \STATE \textbf{Output:} a program whose type unifies with $\tau$
  \STATE \textbf{return} sample$(\mathcal{D}, \theta, \varnothing, \tau)$
\STATE
     \STATE \textbf{function} sample$(\mathcal{D}, \theta, \mathcal{E}, \tau)$:
  \STATE {\bfseries Input:} DSL $\mathcal{D}$, weight vector $\theta$, environment $\mathcal{E}$, type $\tau$
  \STATE \textbf{Output:} a program whose type unifies with $\tau$
  \IF{$\tau = \alpha\to\beta$}
  \STATE var $\gets$ an unused variable name
  \STATE body $\sim$ sample$(\mathcal{D},\theta,\{\text{var}:\alpha\}\cup\mathcal{E},\beta)$
   \STATE \textbf{return} $\lambda \text{var}.$ body
   \ENDIF
   %   \ELSE
   \STATE $\text{primitives} \gets\{p | &p: \tau' \in \mathcal{D}\cup\mathcal{E}$
   \STATE \hspace{2.5cm}$\text{if }\tau\text{ can unify with yield}(\tau') \} $

   \STATE Draw $e\sim \text{primitives}$, w.p. $\propto\theta_e$ if $e\in \mathcal{D}$
   \STATE \hspace{3.1cm}w.p. $\propto\frac{\theta_{var}}{|\text{variables}|}$ if $e\in \mathcal{E}$
   \STATE Let $e:\alpha_1\to\alpha_2\to\cdots\to \alpha_K\to\beta$. Unify $\tau$ with $\beta$.
%   \STATE unify$(\tau,\beta)$
   \FOR{$k=1$ {\bfseries to} $K$}
 \STATE $a_k\sim\text{sample}(\mathcal{D},\theta,\mathcal{E},\alpha_k)$
 \ENDFOR
 \STATE \textbf{return} $e\;a_1\; a_2\; \cdots\; a_K$
 \STATE
 \STATE \textbf{function} yield$(\tau)$:
 \STATE \textbf{Input:} type $\tau$
 \STATE \textbf{Output:} the return type of $\tau$
 \STATE \textbf{if} $\tau = \alpha\to \beta$ \textbf{then return }yield$(\beta)$ \textbf{else return }$\tau$
\end{algorithmic}
\end{algorithm}

@@

The recognition model $q$ is a neural network that predicts,
for each task $x\in X$, a weight vector $q(x) = \theta^{(x)}\in \mathbb{R}^{|\mathcal{D}| + 1}$.
Together with the DSL, this defines a distribution over programs,
$\probability[p|\mathcal{D},\theta = q(x)]$.
We abbreviate this distribution as $q(p|x)$.
The crucial aspect of this framing is that the neural network
can leverage the structure of the DSL,
and is \emph{not} responsible for
generating programs wholesale.
We will show that this lets us get away with a simple, low-capacity neural network.

@@

The purpose of the DSL is to
offer a set of abstractions
that allow an agent to easily express solutions to the tasks at hand.
In the \system algorithm we infer the DSL from a collection of frontiers.
Intuitively, we want the algorithm to
look at the programs in the frontiers and
generalize beyond them; 
not only so the DSL can better express the current solutions,
but also so that the DSL might expose new abstractions
which will later used to
discover even better programs.

@@

Algorithm~\ref{mainAlgorithm} describes how we combine the
program search, recognition model training, and DSL induction.
We additionally make the following implementation details:
(1) On the first iteration we do \emph{not} train the recognition model on samples from
the generative model because, on the first iteration, we have not yet learned a
new DSL, so we instead trained the network to only maximize $\mathcal{L}_{\text{AE}}$;
(2) Because the frontiers can grow very large,
we only keep around the top $10^4$ programs $p$ in each frontier $\mathcal{F}_x$
with the highest $\probability[x,p|\mathcal{D},\theta]$;
(3) During both DSL induction and neural net training, we calculate $\text{score}(\mathcal{D}')$ and $\mathcal{L}_{\text{RM}}$
by only summing over the top $K$
programs in $\mathcal{F}_x$.
We found that $K = 2$ sufficed.

>>>>

The Helmholtz machine~\cite{dayan1995helmholtz} introduced the idea of alternating between
learning a generative model and training a recognition model on samples from the generative model.
We apply this same idea to the problem of program induction,
and so we call our algorithm the \system.
Figure~\ref{feeding} diagrams how these three steps feed off of each other.
\begin{figure}\centering
  \begin{tikzpicture}
    \begin{scope}[shift = {(1,-1)}]
    \node[align = center](DSL) at (6,4) {DSL: $\mathcal{D}$};
    \node[align = center](synthesis) at (4,2) {Search for \\programs: $p$};
    \node[ align = center](recognitionModel) at (6,0) {Recognition \\model: $q$};

@@

\begin{algorithm}[tb]
   \caption{Generative model over programs}
   \label{programGenerativeModel}
   \begin{algorithmic}
     \STATE \textbf{function} sampleProgramFromDSL$(\mathcal{D}, \theta, \tau)$:
  \STATE {\bfseries Input:} DSL $\mathcal{D}$, weight vector $\theta$, type $\tau$
  \STATE \textbf{Output:} a program whose type unifies with $\tau$
  \STATE \textbf{return} sample$(\mathcal{D}, \theta, \varnothing, \tau)$
\STATE
     \STATE \textbf{function} sample$(\mathcal{D}, \theta, \mathcal{E}, \tau)$:
  \STATE {\bfseries Input:} DSL $\mathcal{D}$, weight vector $\theta$, environment $\mathcal{E}$, type $\tau$
  \STATE \textbf{Output:} a program whose type unifies with $\tau$
  \IF{$\tau = \alpha\to\beta$}
  \STATE var $\gets$ an unused variable name
  \STATE body $\sim$ sample$(\mathcal{D},\theta,\{\text{var}:\alpha\}\cup\mathcal{E},\beta)$
   \STATE \textbf{return} \texttt{(lambda (}var\texttt{) }body\texttt{)}
   \ENDIF
   %   \ELSE
   \STATE $\text{primitives} \gets\{p | &p: \tau' \in \mathcal{D}\cup\mathcal{E}$
   \STATE \hspace{2.5cm}$\text{if }\tau\text{ can unify with yield}(\tau') \} $

   \STATE Draw $e\sim \text{primitives}$, w.p. $\propto\theta_e$ if $e\in \mathcal{D}$
   \STATE \hspace{3.1cm}w.p. $\propto\frac{\theta_{var}}{|\text{variables}|}$ if $e\in \mathcal{E}$
   \STATE Unify $\tau$ with yield$(\tau')$.
   \STATE $\left\{\alpha_k \right\}_{k\leq K}\gets\text{argTypes}(\tau')$
%   \STATE unify$(\tau,\beta)$
   \FOR{$k=1$ {\bfseries to} $K$}
 \STATE $a_k\sim\text{sample}(\mathcal{D},\theta,\mathcal{E},\alpha_k)$
 \ENDFOR
 \STATE \textbf{return} \texttt{(}$e\;a_1\; a_2\; \cdots\; a_K$\texttt{)}
 \STATE\textbf{where:}
 \STATE  yield$(\tau) = \begin{cases}
\text{yield}(\beta)   &\text{ if }$\tau = \alpha\to \beta$\\
\tau   &\text{ otherwise.}
 \end{cases}$ 
 \STATE\hspace{0cm} argTypes$(\tau) = \begin{cases}
[\alpha] + \text{argTypes}(\beta)   &\text{ if }$\tau = \alpha\to \beta$\\
[]   &\text{ otherwise.}
 \end{cases}$
\end{algorithmic}
\end{algorithm}

@@

The recognition model $q$ is a neural network that predicts,
for each task $x\in X$, a weight vector $q(x) = \theta^{(x)}\in \mathbb{R}^{|\mathcal{D}| + 1}$.
Together with the DSL, this defines a distribution over programs,
$\probability[p|\mathcal{D},\theta = q(x)]$.
We abbreviate this distribution as $q(p|x)$.
The crucial aspect of this framing is that the neural network
leverages the structure of the learned DSL,
and so is \emph{not} responsible for
generating programs wholesale.

@@

The purpose of the DSL is to
offer a set of abstractions
that allow an agent to easily express solutions to the tasks at hand.
In the \system algorithm we infer the DSL from a collection of frontiers.
Intuitively, we want the algorithm to
look at the programs in the frontiers and
generalize beyond them; 
not only so the DSL can better express the current solutions,
but also so that the DSL might expose new abstractions
which will later be used to
discover even better programs.

@@

Algorithm~\ref{mainAlgorithm} describes how we combine the
program search, recognition model training, and DSL induction.
We additionally make the following implementation details:
(1) On the first iteration we do \emph{not} train the recognition model on samples from
the generative model because, on the first iteration, we have not yet learned a
new DSL, so we instead trained the network to only maximize $\mathcal{L}_{\text{AE}}$;
(2) Because the frontiers can grow very large,
we only keep around the top $10^4$ programs $p$ in each frontier $\mathcal{F}_x$
with the highest $\probability[x,p|\mathcal{D},\theta]$;
(3) During both DSL induction and neural net training, we calculate $\text{score}(\mathcal{D}')$ and $\mathcal{L}_{\text{RM}}$
by only summing over the top $K$
programs in $\mathcal{F}_x$.
We found that $K = 2$ sufficed.
(4) For added robustness, at each iteration we enumerate programs from both the generative model and the recognition model.
